Trap spaces of Boolean networks are conflict-free siphons of their Petri net encoding

### 2023-03-16 manuscripts/TCS2022/tcs2022.tex
Done for this pass…

@@

To evaluate the performance of the newly proposed methods (implemented as a Python package named \texttt{Trappist}) and the state-of-the-art methods (\texttt{bioLQM}\footnote{\url{http://colomoto.org/biolqm/doc/tools-trapspace.html}}, \texttt{PyBoolNet}~\cite{klarner2015computing,klarner2017pyboolnet}, and \texttt{mpbn}~\cite{Paulev2020}), we compared them on both \texttt{PyBoolNet}'s own model repository and many real-world models from various sources in the literature.
To our knowledge, these models are a highly representative sample of Boolean models currently available in the literature.
It is worth noting that \texttt{mpbn}~\cite{Paulev2020} only handles locally-monotonic models, whereas the other methods can handle general models.
To obtain a more comprehensive comparison, we also used random models generated by a third-party software \texttt{BoolNet R} package~\cite{mussel2010boolnet}.
As explained in Section~\ref{sec:case_study}, in our benchmarks, we only searched for the first 1000 minimal trap spaces for each model.
It is worth noting that unlike existing analysis shown in the literature, we did not fix specific values for source nodes in all the considered models.

@@

All the models and a Jupyter notebook realizing the benchmarks can be found at~\url{https://github.com/soli/trap-spaces-as-siphons}.
These can be run on a Docker image in the cloud by clicking the ``Binder'' button.

@@

Table~\ref{tab:pyboolnet_repo} shows the experimental results on the models from the official \texttt{PyBoolNet} repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}.
Column \(n\) denotes the number of nodes of each model.
Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 minimal trap spaces.
``DNF'' means that the method did not finish the computation (stopping at the first 1000 minimal trap spaces) within the time limit of three minutes.
In the case of \texttt{bioLQM}, ``N/A'' means that the number of all minimal trap spaces of the model is larger than 1000 and we did not recorded the running time of \texttt{bioLQM} because it always requires to compute all minimal trap spaces.
A number in bold indicates a ratio greater than three compared to the best result.
``NM'' indicates a non-locally-monotonic model.
There are four variants of \texttt{Trappist}: \texttt{SAT} (i.e., \texttt{Trappist-MaxSAT}, the MaxSAT-based method shown in Subsection~\ref{subsec:computation_csp}), \texttt{CP} (i.e., \texttt{Trappist-CP}, the CP-based method shown in Subsection~\ref{subsec:computation_csp}), \texttt{ILP} (i.e., \texttt{Trappist-ILP}, the ILP-based method shown in Subsection~\ref{subsec:computation_ilp}), and \texttt{ASP} (i.e., \texttt{Trappist-ASP}, the ASP-based method shown in Subsection~\ref{subsec:computation_asp}).

We first analyze the results of the four variants of \texttt{Trappist}.
We can see that \texttt{Trappist-MaxSAT} and \texttt{Trappist-ASP} are comparable in most models, but \texttt{Trappist-ASP} is much faster for the jaoude\_thdiff and selvaggio\_emt models where the number of minimal trap spaces is greater than 1000.
The latter can be explained by the fact that \texttt{Trappist-MaxSAT} follows an iterative approach, i.e., it restarts the search each time a solution is found (see Subsection~\ref{subsec:computation_csp}).
The latter can be explained by the fact that \texttt{Trappist-MaxSAT} follows an iterative approach, i.e., it restarts the search each time a solution is found (see Subsection~\ref{subsec:computation_csp}).
This iterative approach may be less efficient than the way ASP solvers use to enumerate multiple solutions (answer sets), which is an advantage of ASP solvers~\cite{DBLP:journals/aicom/GebserKKOSS11}.
Hence, when the number of solutions increases, the inferiority of \texttt{Trappist-MaxSAT} compared to \texttt{Trappist-ASP} will be exhibited more clearly.
The two remaining variants, \texttt{Trappist-CP} and \texttt{Trappist-ILP}, are much less efficient than \texttt{Trappist-MaxSAT} and \texttt{Trappist-ASP} in every model, even are \(2000 \times\) slower in some models.
The first reason for their bad performance is that they are also iterative methods like \texttt{Trappist-MaxSAT}, thus they are not efficient for "enumeration" problems.
Upon closer inspection, for the Boolean CSP characterizing conflict-free siphons, CP seems to be something that is a "less-efficient-SAT".
For ILP, it may be even worse, since the problem is purely Boolean (no
real or integer numbers whatsoever).
This is confirmed by the observation that for some quite large models (e.g., the grieco\_mapk, zhang\_tlgl, and zhang\_tlgl\_v2 models), \texttt{Trappist-ILP} is much slower than \texttt{Trappist-CP}.
Note that the inferiority of ILP compared to ASP with respect to the trap space enumeration has been reported in~\cite{klarner2015computing}.
Hereafter, we shall compare the best variant of \texttt{Trappist} (i.e., \texttt{Trappist-ASP}) with other methods.

As shown in Table~\ref{tab:pyboolnet_repo}, for most of the models of the \texttt{PyBoolNet} repository, the results are comparable with all minimal trap spaces found very fast.
However upon closer inspection, we can see some notable differences.
First, \texttt{Trappist-ASP} is far more efficient than \texttt{bioLQM} in every model with speedups between \(5\times\) and \(16 \times\).
Second, for small models, \texttt{PyBoolNet} and \texttt{mpbn} are comparable to \texttt{Trappist-ASP}.
However, on every model that was a bit challenging for \texttt{PyBoolNet} or \texttt{mpbn}, \texttt{Trappist-ASP} is far more efficient with speedups between \(3\times\) and \(5 \times\) for the case of \texttt{mpbn}, and between \(5\times\) and \(1783 \times\) for the case of \texttt{PyBoolNet}.
In particular, the second best variant of \texttt{Trappist} (i.e., \texttt{Trappist-MaxSAT}) is even far more efficient than \texttt{bioLQM} and \texttt{PyBoolNet}, is comparable to \texttt{mpbn} on every model.
It is worth noting that for 3 of the 29 models, \texttt{mpbn} did not give any answer because these models are locally-monotonic but all the other methods did, which confirms the limit of \texttt{mpbn} on the applicable class of models.

@@

	\end{semilogxaxis}
	\end{tikzpicture}
	\caption{Cumulative numbers of the \texttt{BBM} models that have less than 1000 minimal trap spaces (above) and \texttt{BBM} models (below) solved by the compared methods with respect to enumerating the first 1000 minimal trap spaces.}
	\label{fig:BBM_cumul}
\end{figure}

Figure~\ref{fig:BBM_cumul} (above) shows cumulative numbers of the \texttt{BBM} models that have less than 1000 minimal trap spaces solved by the compared methods with respect to enumerating the first 1000 minimal trap spaces.
The number of such models is 134 (per all 211 models), and 15 of them are non-locally-monotonic.
This model set allows us to fairly consider \texttt{bioLQM} for comparison, since \texttt{bioLQM} always requires to compute all minimal trap spaces.
We can first see that \texttt{Trappist-ASP} and \texttt{Trappist-MaxSAT} are still the two best methods as they can handle every model within 1s as well as they always can handle more models than all the remaining methods on every time limit.
Second, \texttt{Trappist-CP} is better than \texttt{Trappist-ILP}, which is consistent with their comparison shown in the previous subsection.
Third, one notable remark is that for the time limit of 100s or 180s, \texttt{Trappist-CP} can handle more models than all \texttt{bioLQM}, \texttt{PyBoolNet}, and \texttt{mpbn}.
This remark shows that even with a not best implementation, our alternative approach is still better than the state-of-the-art methods on a certain set of real-world models.
This is supported by the fact that our alternative approach avoids the need for computing prime implicants (as opposed to \texttt{PyBoolNet}) and can handle non-locally-monotonic Boolean networks (as opposed to \texttt{mpbn}).

Figure~\ref{fig:BBM_cumul} (below) shows cumulative numbers of the \texttt{BBM} models solved by the compared methods (except \texttt{bioLQM}, \texttt{Trappist-CP}, and \texttt{Trappist-ILP}) with respect to enumerating the first 1000 minimal trap spaces.
We omit the results of \texttt{Trappist-CP} and \texttt{Trappist-ILP} because they can handle no model with more than 1000 minimal trap spaces.
Again, we can see that \texttt{Trappist-ASP} and \texttt{Trappist-MaxSAT} are the two best methods as they can handle every but one model within 5s as well as they always can handle much more models than both \texttt{PyBoolNet} and \texttt{mpbn} on every time limit.
Note that with the time limit of 0.5s, \texttt{Trappist-ASP} can handle more 14 models than \texttt{Trappist-MaxSAT}, which is opposed to the case of models with less than 1000 minimal trap spaces (see Figure~\ref{fig:BBM_cumul} (above)).
This observation confirms the disadvantage of \texttt{Trappist-MaxSAT} compared to \texttt{Trappist-ASP} for the case of many minimal trap spaces.

@@

First, we obtained some observations on the four variants of \texttt{Trappist} consistent with the observations obtained in the previous subsections.
More specifically, \texttt{Trappist-ASP} is still the best variant with the running time is always less than one second for every model, and following by \texttt{Trappist-MaxSAT}.
In particular, the difference in running time between \texttt{Trappist-ASP} and \texttt{Trappist-MaxSAT} is bigger for larger models or models with more than 1000 minimal trap spaces.
\texttt{Trappist-CP} and \texttt{Trappist-ILP} still have bad performance, with \texttt{Trappist-CP} is a better than \texttt{Trappist-ILP}.
They still can handle no model with more than 1000 minimal trap spaces.
However, \texttt{Trappist-CP} or \texttt{Trappist-ILP} can handle the FT-GRN and Pluripotency models, whereas all \texttt{bioLQM}, \texttt{PyBoolNet}, and \texttt{mpbn} cannot.

@@

Third, for 11 of the 32 models (more than 34\%), \texttt{mpbn} did not give any answer because these models as non-locally-monotonic.
For 21 of the 32 models where \texttt{mpbn} returned the answers, \texttt{mpbn} and \texttt{Trappist-ASP} are comparable in computation time, with \texttt{mpbn} appears quite slower on average.
In particular, for the Rho-GTPases model, \texttt{mpbn} is \(577\times\) slower than \texttt{Trappist-ASP}.
This observation along with the comparisons between \texttt{mpbn} and \texttt{Trappist-ASP} in the previous subsections are quite surprising because the ASP encoding of \texttt{mpbn} only requires the DNF for the activation part of a Boolean function, whereas that of \texttt{Trappist-ASP} requires both the activation and inhibition parts (see Subsection~\ref{subsec:computation_asp}).
However, the reason maybe lies on the differences in the ASP encoding characteristics of the two methods and the fact that \texttt{mpbn} needs to spend time for checking the locally-monotonicity of each Boolean function in a Boolean network.
It is possible that \texttt{mpbn} may outperform \texttt{Trappist} for a certain set of models, but not for the set of real-world models considered in this article.

@@

We randomly generated a set of N-K models~\cite{glass1973logical} with network size \(n\) in the set \{100, 150, 200, 250, 300, 350, 400\} and \(K = 3\) (i.e., each node has exactly three input nodes).
We chose N-K models because they are a useful tool for studying the dynamics of Boolean networks~\cite{glass1973logical,klarner2015computing,Rozum2021}.
For each network size, 50 instances were generated using the \verb|generateRandomNKNetwork| function.
In total, we have 350 random models.
We then applied the compared methods to these models and recorded the running time of each method for each model.
It is worth noting that N-K models usually have small numbers of minimal trap spaces~\cite{klarner2015computing}.
Hence, we searched for all solutions in each model, which makes the comparison to \texttt{bioLQM} more comprehensive.
In addition, each node has only three input nodes, i.e., the number of prime-implicants of the associated Boolean function is small.
Hence, \texttt{PyBoolNet} always passed the phase of computing prime-implicants in every model even within one second, which enables us to compare the ASP encoding of \texttt{PyBoolNet} and that of \texttt{Trappist-ASP}.

@@

Figure~\ref{fig:random_cumul_first} shows cumulative numbers of random models solved by the compared methods with respect to enumerating all the minimal trap spaces.
The number of succeeded models within three minutes for each method is: \texttt{bioLQM} (0), \texttt{PyBoolNet} (320), \texttt{mpbn} (0), \texttt{Trappist-maxSAT} (338), \texttt{Trappist-CP} (226), \texttt{Trappist-ILP} (39), \texttt{Trappist-ASP} (349).
We can see that \texttt{Trappist-ASP} is the only method that can handle every but one model.
Note that none of the other methods can handle the only model failed by \texttt{Trappist-ASP}.
We also obtained some observations consistent with those obtained for real-world models.
More specifically, \texttt{Trappist-MaxSAT} is still the second best method and \texttt{Trappist-CP} is better than \texttt{Trappist-ILP}.
Upon closer inspection, we obtained several notable observations as follows.

First, \texttt{mpbn} did not be able to handle any model because all the models are non-locally-monotonic.
Recall that a Boolean network is non-locally-monotonic if only one of its Boolean functions is non-locally-monotonic.
Hence, it is apparent that all the randomly generated models are non-locally-monotonic because of the number of nodes is large (\(n \geq 100\)).
This observation confirms the limit on the applicable model class of \texttt{mpbn}.

@@

Third, for every time limit, \texttt{Trappist-ASP} can always handle much models than \texttt{PyBoolNet}, ranging from 29 to 65 more models.
Since the time for the phase of computing prime-implicants of \texttt{PyBoolNet} is negligible in every model, most of the running time of \texttt{PyBoolNet} was spent for its ASP solving phase.
Hence, we can easily see that the ASP encoding of \texttt{Trappist-ASP} is much better than that of \texttt{PyBoolNet}. 
This observation is consistent with the theoretical comparison in the ASP encoding between \texttt{Trappist-ASP} and \texttt{PyBoolNet} mentioned in Subsection~\ref{subsec:computation_asp}.

@@

We have tested our alternative approach on many Boolean network models of various sizes and types (e.g., real-world models, randomly generated models).
This indicates the high coverage and comprehensiveness of the experiments.

Among the four variants of the alternative approach, \texttt{Trappist-ASP} is the best method as it vastly outperforms all the other variants.
The second best one is \texttt{Trappist-MaxSAT}.
The two remaining variants (i.e., \texttt{Trappist-CP} and \texttt{Trappist-ILP}) give bad performance for most models.
However, for some certain cases, they are even better than all the state-of-the-art methods (i.e., \texttt{bioLQM}, \texttt{PyBoolNet}, and \texttt{mpbn}).
This is evidence for the advantages of the alternative approach compared to the state-of-the-art ones.

Regarding general Boolean networks, \texttt{Trappist-ASP} (even \texttt{Trappist-MaxSAT}) is far more efficient than both \texttt{bioLQM} and \texttt{PyBoolNet}.
The speedups of \texttt{Trappist-ASP} or \texttt{Trappist-MaxSAT} are large, even between one and three orders of magnitude for most models.
In addition, the experimental results also confirm that the ASP encoding of \texttt{Trappist-ASP} is much better than that of \texttt{PyBoolNet}.

Regarding locally-monotonic Boolean networks, the performance of \texttt{mpbn} is comparable to that of \texttt{Trappist-ASP} or \texttt{Trappist-MaxSAT}.
However, \texttt{mpbn} is quite slower than \texttt{Trappist-ASP} in average.
This shows the practical advantage of \texttt{Trappist-ASP} compared to \texttt{mpbn}, though its ASP encoding may be more complex than that of \texttt{mpbn} in theory.

@@

In this article we have explored and proved for the first time the equivalence between (minimal) trap spaces of a general Boolean network and (maximal) conflict-free siphons of its Petri net encoding.
We have shown several important applications of this finding to studying properties of trap spaces in Boolean networks.
As an important practical application of the equivalence, we have proposed a new approach for the computation of minimal trap spaces in Boolean networks, based on the enumeration of maximal conflict-free siphons of Petri nets.
We have also proposed the four possible methods using MaxSAT, CP, ILP, and ASP for implementing the new approach.
The proposed methods have been evaluated on many real-world models from the literature as well as randomly generated models.
The experimental results show that the new approach vastly outperforms all the state-of-the-art methods in terms of general Boolean networks and is comparable to the \texttt{mpbn} method even much better in average in terms of locally-monotonic Boolean networks.
We believe that this opens up the way to a much better analysis of large Boolean networks, which is needed with the advent of automatic model-generation pipelines~\cite{ostaszewski2021covid19}.

@@

>>>>

\sylvain{I think that the result above (0.19s) is only about the F5 model, but do we also have the time for the F4 model, if so, it should be nice to include it too.}

@@

To evaluate the performance of the newly proposed methods (implemented as a Python package named \texttt{Trappist} and available on the Python package index\footnote{\url{https://pypi.org/project/trappist/}}) and the state-of-the-art methods (\texttt{bioLQM}\footnote{\url{http://colomoto.org/biolqm/doc/tools-trapspace.html}}, \texttt{PyBoolNet}~\cite{klarner2015computing,klarner2017pyboolnet}, and \texttt{mpbn}~\cite{Paulev2020}), we compared them on both \texttt{PyBoolNet}'s own model repository and many real-world models from various sources in the literature.
To our knowledge, these models are a highly representative sample of Boolean models currently available.
It is worth noting that \texttt{mpbn}~\cite{Paulev2020} only handles locally-monotonic models, whereas the other methods can handle general models.
To obtain a more comprehensive comparison, we also used random models generated by a third-party software \texttt{BoolNet R} package~\cite{mussel2010boolnet}.
As explained in Section~\ref{sec:case_study}, in our benchmarks, we only searched for the first 1000 minimal trap spaces for each model.
It is worth noting that unlike existing analysis shown in the literature, we did not fix specific values for source nodes in all the considered models.

@@

All the models and some Jupyter notebooks realizing the benchmarks (and named \texttt{TCS-Benchmark-<...>.ipynb}) can be found at~\url{https://github.com/soli/trap-spaces-as-siphons/}.
These can be run on a Docker image in the cloud by clicking the ``Binder'' button.

@@

Table~\ref{tab:pyboolnet_repo} shows the experimental results on the models from the official \texttt{PyBoolNet} repository\footnote{\url{https://github.com/hklarner/pyboolnet/tree/master/pyboolnet/repository}}.
Column \(n\) denotes the number of nodes of each model.
Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 minimal trap spaces.
``DNF'' means that the method did not finish the computation within the time limit of three minutes.
In the case of \texttt{bioLQM}, ``N/A'' means that the number of all minimal trap spaces of the model is larger than 1000 and we did not recorded the running time of \texttt{bioLQM} because it always requires to compute all minimal trap spaces.
A number in bold indicates a ratio greater than three compared to the best result.
``NM'' indicates a non-locally-monotonic model.
\sylvain{should we also highlight in some way the (tied-)best result?}
There are four variants of \texttt{Trappist}: \texttt{SAT} (i.e., \texttt{Trappist-MaxSAT}, the MaxSAT-based method shown in Subsection~\ref{subsec:computation_csp}), \texttt{CP} (i.e., \texttt{Trappist-CP}, the CP-based method shown in Subsection~\ref{subsec:computation_csp}), \texttt{ILP} (i.e., \texttt{Trappist-ILP}, the ILP-based method shown in Subsection~\ref{subsec:computation_ilp}), and \texttt{ASP} (i.e., \texttt{Trappist-ASP}, the ASP-based method shown in Subsection~\ref{subsec:computation_asp}).

We first analyze the results of the four variants of \texttt{Trappist}.
We can see that \texttt{Trappist-MaxSAT} and \texttt{Trappist-ASP} are comparable in most models, but \texttt{Trappist-ASP} is much faster for the jaoude\_thdiff and selvaggio\_emt models where the number of minimal trap spaces is greater than 1000.
The latter can be explained by the fact that \texttt{Trappist-MaxSAT} follows an iterative approach, i.e., it restarts the search with a new constraint each time a solution is found (see Subsection~\ref{subsec:computation_csp}).
This iterative approach may be less efficient than the way ASP solvers use to enumerate multiple solutions (answer sets), which is an advantage of ASP solvers~\cite{DBLP:journals/aicom/GebserKKOSS11}.
Hence, when the number of solutions increases, the inferiority of \texttt{Trappist-MaxSAT} compared to \texttt{Trappist-ASP} will be exhibited more clearly.
The two remaining variants, \texttt{Trappist-CP} and \texttt{Trappist-ILP}, are much less efficient than \texttt{Trappist-MaxSAT} and \texttt{Trappist-ASP} in every model, even are more than three orders of magnitude slower in some models.
The first reason for their bad performance is that they are also iterative methods like \texttt{Trappist-MaxSAT}, thus they are not efficient for "enumeration" problems.
Upon closer inspection, for the Boolean CSP characterizing conflict-free siphons, CP seems to be something that is a "less-efficient-SAT", handling mostly Boolean constraints and making little use of the global constraints only added for the iterative part.
For ILP, it may be even worse, since the problem is purely Boolean (no
real or integer numbers whatsoever).
This is confirmed by the observation that for some quite large models (e.g., the grieco\_mapk, zhang\_tlgl, and zhang\_tlgl\_v2 models), \texttt{Trappist-ILP} is much slower than \texttt{Trappist-CP}.
Note that the inferiority of ILP compared to ASP with respect to the trap space enumeration has been reported in~\cite{klarner2015computing}.
Hereafter, we shall compare the best variant of \texttt{Trappist} (i.e., \texttt{Trappist-ASP}) with other methods.

As shown in Table~\ref{tab:pyboolnet_repo}, for most of the models of the \texttt{PyBoolNet} repository, the results are comparable with all minimal trap spaces found very fast.
However upon closer inspection, we can see some notable differences.
First, \texttt{Trappist-ASP} is far more efficient than \texttt{bioLQM} in every model with speedups between \(5\times\) and \(16 \times\).
Second, for small models, \texttt{PyBoolNet} and \texttt{mpbn} are comparable to \texttt{Trappist-ASP}.
However, on every model that was a bit challenging for \texttt{PyBoolNet} or \texttt{mpbn}, \texttt{Trappist-ASP} is far more efficient with speedups between \(3\times\) and \(5 \times\) for the case of \texttt{mpbn}, and between \(5\times\) and \(1783 \times\) for the case of \texttt{PyBoolNet}.
In particular, the second best variant of \texttt{Trappist} (i.e., \texttt{Trappist-MaxSAT}) is even far more efficient than \texttt{bioLQM} and \texttt{PyBoolNet}, and is comparable to \texttt{mpbn} on every model.
It is worth noting that for 3 of the 29 models, \texttt{mpbn} did not give any answer because these models are locally-monotonic but all the other methods did, which confirms the limit of \texttt{mpbn} on the applicable class of models.

@@

	\end{semilogxaxis}
	\end{tikzpicture}
	\caption{Cumulative numbers of the \texttt{BBM} models that have less than 1000 minimal trap spaces (upper panel) and \texttt{BBM} models solved by the compared methods with respect to enumerating the first 1000 minimal trap spaces (lower panel).}
	\label{fig:BBM_cumul}
\end{figure}

Figure~\ref{fig:BBM_cumul} (upper panel) shows cumulative numbers of the \texttt{BBM} models that have less than 1000 minimal trap spaces solved by the compared methods with respect to enumerating the first 1000 minimal trap spaces.
The number of such models is 134 (per all 211 models), and 15 of them are non-locally-monotonic.
This model set allows us to fairly consider \texttt{bioLQM} for comparison, since \texttt{bioLQM} always requires to compute all minimal trap spaces.
We can first see that \texttt{Trappist-ASP} and \texttt{Trappist-MaxSAT} are still the two best methods as they can handle every model within 1s as well as they always can handle more models than all the remaining methods on every time limit.
Second, \texttt{Trappist-CP} is better than \texttt{Trappist-ILP}, which is consistent with their comparison shown in the previous subsection.
Third, one notable remark is that for the time limit of 100s or 180s, \texttt{Trappist-CP} can handle more models than all \texttt{bioLQM}, \texttt{PyBoolNet}, and \texttt{mpbn}.
This remark shows that even with a not best implementation, our alternative approach is still better than the state-of-the-art methods on a certain set of real-world models.
This is supported by the fact that our alternative approach avoids the need for computing prime implicants (as opposed to \texttt{PyBoolNet}) and can handle non-locally-monotonic Boolean networks (as opposed to \texttt{mpbn}).

Figure~\ref{fig:BBM_cumul} (lower panel) shows cumulative numbers of the \texttt{BBM} models solved by the compared methods (except \texttt{bioLQM}, \texttt{Trappist-CP}, and \texttt{Trappist-ILP}) with respect to enumerating the first 1000 minimal trap spaces.
We omit the results of \texttt{Trappist-CP} and \texttt{Trappist-ILP} because they can handle no model with more than 1000 minimal trap spaces.
Again, we can see that \texttt{Trappist-ASP} and \texttt{Trappist-MaxSAT} are the two best methods as they can handle every but one model within 5s. They also always handle many more models than both \texttt{PyBoolNet} and \texttt{mpbn} on every time limit.
Note that with the time limit of 0.5s, \texttt{Trappist-ASP} can handle 14 more models than \texttt{Trappist-MaxSAT}, which is opposed to the case of models with less than 1000 minimal trap spaces (see Figure~\ref{fig:BBM_cumul} (upper panel)).
This observation confirms the disadvantage of \texttt{Trappist-MaxSAT} compared to \texttt{Trappist-ASP} for the case of many minimal trap spaces.

@@

First, we obtained some observations on the four variants of \texttt{Trappist} consistent with the observations obtained in the previous subsections.
More specifically, \texttt{Trappist-ASP} is still the best variant with a running time below one second for every model, and followed by \texttt{Trappist-MaxSAT}.
In particular, the difference in running time between \texttt{Trappist-ASP} and \texttt{Trappist-MaxSAT} is bigger for larger models or models with more than 1000 minimal trap spaces.
\texttt{Trappist-CP} and \texttt{Trappist-ILP} still have a much worse performance, with \texttt{Trappist-CP} better than \texttt{Trappist-ILP}.
They still can handle no model with more than 1000 minimal trap spaces.
However, \texttt{Trappist-CP} or \texttt{Trappist-ILP} can handle the FT-GRN and Pluripotency models, whereas all \texttt{bioLQM}, \texttt{PyBoolNet}, and \texttt{mpbn} cannot.

@@

Third, for 11 of the 32 models (more than 34\%), \texttt{mpbn} did not give any answer because these models as non-locally-monotonic.
For 21 of the 32 models where \texttt{mpbn} returned the answers, \texttt{mpbn} and \texttt{Trappist-ASP} are roughly comparable in computation time, but \texttt{mpbn} appears quite slower on average.
In particular, for the Rho-GTPases model, \texttt{mpbn} is \(577\times\) slower than \texttt{Trappist-ASP}.
This observation along with the comparisons between \texttt{mpbn} and \texttt{Trappist-ASP} in the previous subsections are quite surprising because the ASP encoding of \texttt{mpbn} only requires the DNF for the activation part of a Boolean function, whereas that of \texttt{Trappist-ASP} requires both the activation and inhibition parts (see Subsection~\ref{subsec:computation_asp}).
However, the reason may lie on the differences in the ASP encoding characteristics of the two methods and the fact that \texttt{mpbn} needs to spend time checking the locally-monotonicity of each Boolean function in a Boolean network.
We expect that \texttt{mpbn} may outperform \texttt{Trappist} for a certain set of models, but not for the set of real-world models considered in this article.

@@

We randomly generated a set of N-K models~\cite{glass1973logical} with network size \(n\) in the set \{100, 150, 200, 250, 300, 350, 400\} and in-degree \(K = 3\) (i.e., each node has exactly three input nodes).
We chose N-K models because they are a useful tool for studying the dynamics of Boolean networks~\cite{glass1973logical,klarner2015computing,Rozum2021}.
For each network size, 50 instances were generated using the \verb|generateRandomNKNetwork| function.
In total, we have 350 random models.
We then applied the compared methods to these models and recorded the running time of each method for each model.
It is worth noting that N-K models usually have small numbers of minimal trap spaces~\cite{klarner2015computing}.
Hence, we searched for all solutions in each model, which makes the comparison to \texttt{bioLQM} more comprehensive.
In addition, each node has only three input nodes, i.e., the number of prime-implicants of the associated Boolean function is small.
Hence, \texttt{PyBoolNet} always passed the phase of computing prime-implicants in every model even within one second, which enables us to compare the ASP encoding of \texttt{PyBoolNet} and that of \texttt{Trappist-ASP}.

@@

Figure~\ref{fig:random_cumul_first} shows cumulative numbers of random models solved by the compared methods with respect to enumerating all the minimal trap spaces.
The number of succeeded models within three minutes for each method is: \texttt{bioLQM} (0), \texttt{PyBoolNet} (320), \texttt{mpbn} (0), \texttt{Trappist-maxSAT} (338), \texttt{Trappist-CP} (226), \texttt{Trappist-ILP} (39), \texttt{Trappist-ASP} (349).
We can see that \texttt{Trappist-ASP} is the only method that can handle every model, but one.
Note that none of the other methods can handle that only model failed by \texttt{Trappist-ASP}.
We also obtained some observations consistent with those obtained for real-world models.
More specifically, \texttt{Trappist-MaxSAT} is still the second best method and \texttt{Trappist-CP} is better than \texttt{Trappist-ILP}.
Upon closer inspection, we obtained several notable observations as follows.

First, \texttt{mpbn} was not able to handle any model because all the models are non-locally-monotonic.
Recall that a Boolean network is non-locally-monotonic if only one of its Boolean functions is non-locally-monotonic.
Hence, it is apparent that all this type of randomly generated models are non-locally-monotonic because of the number of nodes is large (\(n \geq 100\)).
This observation confirms a limit on the applicable model class of \texttt{mpbn}.

@@

Third, for every time limit, \texttt{Trappist-ASP} can always handle many more models than \texttt{PyBoolNet}, ranging from 29 to 65 more models.
Since the time for the phase of computing prime-implicants of \texttt{PyBoolNet} is negligible in every model, most of the running time of \texttt{PyBoolNet} was spent for its ASP solving phase.
Hence, we can easily see that the ASP encoding of \texttt{Trappist-ASP} is much better than that of \texttt{PyBoolNet}. 
This observation is consistent with the theoretical comparison in the ASP encoding between \texttt{Trappist-ASP} and \texttt{PyBoolNet} mentioned in Subsection~\ref{subsec:computation_asp}.

@@

We have tested our alternative approach on many Boolean network models of various sizes and types (e.g., real-world models, randomly generated models) on existing and newly created benchmarks.
This indicates the high coverage and comprehensiveness of the experiments.

Among the four variants of the alternative approach, \texttt{Trappist-ASP} is the best method as it vastly outperforms all the other variants.
The second best one is \texttt{Trappist-MaxSAT}.
The two remaining variants (i.e., \texttt{Trappist-CP} and \texttt{Trappist-ILP}) give bad performance for most models.
However, for certain cases, they are still better than all state-of-the-art methods (i.e., \texttt{bioLQM}, \texttt{PyBoolNet}, and \texttt{mpbn}).
This is evidence for the advantages of an alternative approach compared to what preexisted.

Regarding general Boolean networks, \texttt{Trappist-ASP} (even \texttt{Trappist-MaxSAT}) is far more efficient than both \texttt{bioLQM} and \texttt{PyBoolNet}.
The speedups of \texttt{Trappist-ASP} or \texttt{Trappist-MaxSAT} are large, even between one and three orders of magnitude for most models.
In addition, the experimental results also confirm that the ASP encoding of \texttt{Trappist-ASP} is much more efficient than that of \texttt{PyBoolNet}.

Regarding locally-monotonic Boolean networks, the performance of \texttt{mpbn} is roughly comparable to that of \texttt{Trappist-ASP} or \texttt{Trappist-MaxSAT}.
However, \texttt{mpbn} is quite slower than \texttt{Trappist-ASP} on average.
This shows the practical advantage of \texttt{Trappist-ASP} compared to \texttt{mpbn}, though its ASP encoding may be more complex than that of \texttt{mpbn} in theory.

@@

In this article we have explored and proved for the first time the equivalence between (minimal) trap spaces of a general Boolean network and (maximal) conflict-free siphons of its Petri net encoding.
We have shown several important applications of this finding to studying properties of trap spaces in Boolean networks.
As an important practical application of the equivalence, we have proposed a new approach for the computation of minimal trap spaces in Boolean networks, based on the enumeration of maximal conflict-free siphons of Petri nets.
We have also proposed the four possible methods using MaxSAT, CP, ILP, and ASP for implementing the new approach.
The proposed methods have been evaluated on many real-world models from the literature as well as randomly generated models.
The experimental results show that the new approach vastly outperforms all the state-of-the-art methods in terms of general Boolean networks and is comparable to the \texttt{mpbn} method even much better on average in terms of locally-monotonic Boolean networks.
We believe that this opens up the way to a much better analysis of large Boolean networks, which is needed with the advent of automatic model-generation pipelines~\cite{ostaszewski2021covid19}.

@@

\sylvain{Do you want to say a word about specific types of trap-spaces? This currently does not really appear in the conclusion.}
